---
title: "NFL Distance Analysis"
format: pdf
editor: visual
---

## Part 1: Overview

This document analyzes the effect of NFL team's travel distance on game performance by analyzing different metrics.

## Background

NFL teams are based on geographic areas and have a home stadium/facility that they train from. Teams play an 18 week regular season, and have another 4 weeks of playoffs depending on regular season performance. Each team plays every weekend, so there becomes a lot of travel involved around the weekend when teams play. When traveling for away games, some teams can travel up to 3,000 miles going from coast to coast. Therefore there may be a relationship between exhausting long distance travel and game day performance. To analyze the effects of NFL team's travel, we look at not only win percentage, but more individual statistics such as defensive interception rate, offensive rush yard averages, and offensive 3rd down conversion rates for example.

------------------------------------------------------------------------

## Challenges

One challenge we encountered was getting the distances that NFL team's had to travel. Unlike the statistics for individual games, the travel distance of teams was not already in an established data package to import. To get the travel distance, we first manually got the rough latitude and longitude coordinates for every team's stadium, then mapped the coordinates to a team's acronym. Then by importing the geosphere package we were able to get the straight line distance between two stadiums in meters. Finally converting this distance into miles gives our metric.

------------------------------------------------------------------------

## Part 2: The initial stats

### Overview

In this segment we will be looking at the base performance stats for the away teams not putting the specific distance into consideration yet.

### Win/Lose percentage

The first metric we decided to look at was the base win/lose percentage as even though this won't tell the full story, it would be a good starting point.

```{r}

#load packages
library(nflreadr)
library(dplyr)

#load schedule data (example: multiple seasons)
games <- load_schedules(2022:2024)

#calculates away win percentage
away_win_pct <- games %>%
  filter(!is.na(home_score), !is.na(away_score)) %>%  #removes games without scores
  mutate(away_win = away_score > home_score) %>%
  summarise(
    total_games = n(),
    away_wins = sum(away_win),
    away_win_percentage = away_wins / total_games
  )
away_win_pct

```

As you can see from the data above during the 2022-2024 seasons, the win/lose rates for the away teams was about a 44%. This may not seem significant as 44% is close to 50% but these stats are similar throughout every single season. It's not like some seasons its above 50% and some are below, they are all consistently close to the 44%. This gives us reason to believe that there may be something going on where the away side have some sort of disadvantaged. Before we continue we will look at some more specific stats for these teams using the first 8 teams for room.

### Team Compare Table

```{r}

# Load libraries
library(nflreadr)
library(dplyr)
library(tidyr)
library(knitr)
library(ggplot2)

# Load data
games <- load_schedules(2023)
pbp <- load_pbp(2023)

# Step 1: Aggregate stats at team-game level
team_game_stats <- pbp %>%
  filter(!is.na(posteam)) %>%
  group_by(game_id, team = posteam) %>%
  summarise(
    completion_pct = mean(complete_pass, na.rm = TRUE),
    total_yards = sum(yards_gained, na.rm = TRUE),
    avg_yards = mean(yards_gained, na.rm = TRUE),
    total_epa = sum(epa, na.rm = TRUE),
    interceptions = sum(interception, na.rm = TRUE),
    .groups = "drop"
  )

# Step 2: Label games as home or away
team_games <- games %>%
  select(game_id, home_team, away_team) %>%
  pivot_longer(
    cols = c(home_team, away_team),
    names_to = "location",
    values_to = "team"
  ) %>%
  mutate(location = ifelse(location == "home_team", "home", "away"))

# Step 3: Join stats with location
team_stats_with_location <- team_games %>%
  left_join(team_game_stats, by = c("game_id", "team"))

# Step 4: Compute averages
team_home_away_split <- team_stats_with_location %>%
  group_by(team, location) %>%
  summarise(
    avg_completion_pct = mean(completion_pct, na.rm = TRUE),
    avg_total_yards = mean(total_yards, na.rm = TRUE),
    avg_epa = mean(total_epa, na.rm = TRUE),
    avg_interceptions = mean(interceptions, na.rm = TRUE),
    games = n(),
    .groups = "drop"
  )

```

```{r}

completion_table <- team_home_away_split %>%
  select(team, location, avg_completion_pct) %>%
  pivot_wider(names_from = location, values_from = avg_completion_pct)

kable(head(completion_table, 8), digits = 2,
      caption = "Completion Percentage (Home vs Away)")

```

```{r}

yards_table <- team_home_away_split %>%
  select(team, location, avg_total_yards) %>%
  pivot_wider(names_from = location, values_from = avg_total_yards)

kable(head(yards_table, 8), digits = 1,
      caption = "Average Total Yards (Home vs Away)")

```

```{r}

epa_table <- team_home_away_split %>%
  select(team, location, avg_epa) %>%
  pivot_wider(names_from = location, values_from = avg_epa)

kable(head(epa_table, 8), digits = 2,
      caption = "Average EPA (Home vs Away)")

```

```{r}

int_table <- team_home_away_split %>%
  select(team, location, avg_interceptions) %>%
  pivot_wider(names_from = location, values_from = avg_interceptions)

kable(head(int_table, 8), digits = 2,
      caption = "Average Interceptions (Home vs Away)")

```
